---
title: "Assignment2SDS315"
author: "Kapil Taspa"
date: "2025-01-28"
output: pdf_document
---

***UTEID**:* kt27955

***Github Repo:*** <https://github.com/ktaspa/Homework2SDS315>

# Problem 1

```{r, echo = FALSE, message = FALSE}
profs <- read.csv('profs.csv')
library(ggplot2)
```

### Part A

```{r, echo = FALSE}
ggplot(profs, aes(x = eval)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "red") +
  labs(
    title = "Distribution of Course Evaluation Scores",
    x = "Evaluation Score",
    y = "Count",
    caption = 'This histogram of the scores shows that the majority of course evaluations  were in the 3.5-4.5 range.')
```

### Part B

```{r, echo = FALSE}
ggplot(profs, aes(x = native, y = eval, fill = native)) +
  geom_boxplot() +
  scale_fill_manual(values = c("blue", "red")) + 
  labs(
    title = "Course Evaluation scores grouped by native English Speaker Status",
    x = "Native English Speaker",
    y = "Evaluation Score",
    caption = 'The distributions show that native english speakers had higher scores than those who werent')
```

### Part C

```{r, echo = FALSE}
ggplot(profs, aes(x = eval)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "red") +
  facet_wrap(~gender,nrow =2) +
  labs(
    title = "Distribution of Course Evaluation Scores by gender",
    x = "Evaluation Score",
    y = "Count",
    caption = 'The distributions of male and female course evaluations appear to be approximately similar when comparing them visually.'
  )
```

### Part D

```{r, echo = FALSE}
ggplot(profs, aes(x = beauty, y = eval)) +
  geom_point() +
  labs(
    title = "Association Between Prof. Beauty and Course Evaluations",
    x = "Physical Attractiveness (beauty rating)",
    y = "Course Evaluation Score",
    caption = 'There appears to be no relationship between professor beauty and evaluation score. 
                Although there is a gap in the low physical attractiveness, high evaluation score area,
                there appears to be no association between the two.'
  )
```

# Problem 2

```{r, echo = FALSE, message = FALSE}
bikeshare <- read.csv('bikeshare.csv')
library(dplyr)
```

### Plot 1

```{r, echo = FALSE}
avg <- bikeshare %>%
  group_by(hr) %>%
  summarize(mean_total = mean(total, na.rm = TRUE))

ggplot(avg, aes(x =hr, y=mean_total)) +
  geom_line() +
  labs(
    title = "Average Hourly Bike Rentals",
    x = "Hour of the Day",
    y = "Average number of bike rentals",
    caption = "THe line graph shows the daily average of total bike rentals by hour across all days in 2011–2012. 
              There appears to be higher rentals during commuting hours or rush hours as we can see 2 peaks around
              the 8 and 17 hour marks. We also see a higher peak in the evening compared to the morning times."
  )
```

### Plot 2

```{r, echo = FALSE}
avgworking <- bikeshare %>%
  group_by(hr,workingday) %>%
  summarize(mean_total = mean(total, na.rm = TRUE), .groups = "drop")

ggplot(avgworking, aes(x =hr, y =mean_total)) +
  geom_line() +
  facet_wrap(~workingday, ncol = 1) +
  labs(
    title = "Average Hourly Bike Rentals by Working Day Status",
    x = "Hour of the Day (0–23)",
    y = "Average Bike Rentals (count)",
    caption = "This facetted line graph compares average hourly rentals on working days (top) vs. non working days (bottom).
               There appears to be a morning and evening peak on working days which makes sense because of commuting times.
                We can also notice that there are less bike rentals overall on non working days because there is no commuting."
  )
```

### Plot 3

```{r, echo = FALSE}
bikesharesubset <- bikeshare %>%
  filter(hr == 9) %>%
  group_by(weathersit, workingday) %>%
  summarize(mean_total = mean(total, na.rm = TRUE), .groups = "drop")

ggplot(bikesharesubset, aes(x = factor(weathersit), y = mean_total)) +
  geom_bar(stat = "identity", color = 'black', fill = 'blue') +
  facet_wrap(~workingday) +
  labs(
    title = "Average 9 AM Bike Rentals by Weather Situation and Working Day Status",
    x = "Weather Situation (1=Clear, 2=Mist/Cloudy, 3=Light Snow/Rain)",
    y = "Average Bike Rentals at 9 AM",
    caption = "This plot plot shows the average 9 am bike rentals grouped by weather situation and
              working day status where 0 and 1 represent non working and working days and 1, 2, and
              3 represent clear, cloudy/mist, and light snow/rain respectively. We see that in all
              weather, there is always moer bike rentals on working days at 9am then non working days.
              In addition, there is always the most bike rentals at 9 am in clear weather."
  )
```

# Problem 3

```{r, echo = FALSE, message = FALSE}
capmetro <- read.csv('capmetro_UT.csv')

capmetro <- mutate(capmetro,
  day_of_week = factor(day_of_week, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
  month = factor(month, levels = c("Sep", "Oct", "Nov"))
)
```

### Plot 1

***Does the hour of peak boardings change from day to day, or is it broadly similar across days?***\
The hour of peak boardings is broadly similar across days.

***Why do you think average boardings on Mondays in September look lower, compared to other days and months?***\
Labor day is in september which means there is a lot less boardings on that Monday which would decrease the average for the month.

Similarly, why do you think average boardings on Weds/Thurs/Fri in November look lower\
Many students are leaving for the holidays during these times especially around the Thanksigiving break time.

```{r, echo = FALSE, warning = FALSE}
boarding_summary <- capmetro %>%
  group_by(day_of_week, hour_of_day, month) %>%
  summarize(avg_boarding = mean(boarding, na.rm = TRUE), .groups = "drop")

ggplot(boarding_summary, aes(x =hour_of_day, y =avg_boarding, color =month, group =month)) +
  geom_line(size = 1) +
  facet_wrap(~day_of_week) +
  labs(
    title = "Average Hourly Boardings by Day of the Week and Month",
    x = "Hour of the Day (6am-10pm)",
    y = "Average Boardings",
    color = "Month",
    caption = "This faceted line graph shows the average number of boardings
              per hour across different days of the week. Peak ridership appears
              to be in the morning and afternoon."
  )
```

### Plot 2

Temperature does not seem to have a strong effect or ridership.

```{r, echo = FALSE}
capmetro_ut <- capmetro %>%
  mutate(weekend = factor(weekend, levels = c(0, 1), labels = c("Weekday", "Weekend")))

ggplot(capmetro_ut, aes(x = temperature, y = boarding, color = weekend)) +
  geom_point() +  # Add transparency for better visibility
  facet_wrap(~hour_of_day, ncol = 3) +  # Facet by hour of the day
  labs(
    title = "Effect of Temperature on UT Bus Ridership, by Hour",
    x = "Temperature (°F)",
    y = "Boardings",
    color = "Day Type",
    caption = "This scatter plot shows the association between temperature and boardings across different hours. 
               There isnt strong correlation between temperature and boardings which means students rely on buses 
               regardless of weather conditions. Higher variation is observed during peak commuting hours."
  )
```

# Problem 4

```{r, echo = FALSE, message = FALSE}
billboard <- read.csv('billboard.csv')
```

### Part A

The table below shows the top 10 most popular songs since 1958 on the top 100 billboard. The table has 3 columns performer, song, and total_weeks. The total weeks that these songs were on the billboard range from 64 to 87.

```{r, echo = FALSE}
billboard %>%
  group_by(performer, song) %>%
  summarize(total_weeks = n(), .groups = "drop") %>%
  arrange(desc(total_weeks)) %>%
  slice_head(n = 10)
```

### Part B

```{r, echo = FALSE}
diversity <- billboard %>%
  filter(year > 1958 & year < 2021) %>% 
  group_by(year) %>%
  summarize(unique_songs = n_distinct(song), .groups = "drop")

ggplot(diversity, aes(x =year, y =unique_songs)) +
  geom_line() +
  labs(
    title = "Musical Diversity on Billboard 100 from 1958-2021",
    x = "Year",
    y = "Number of Unique Songs",
    caption = "This line graph shows how the number of unique songs appearing on the Billboard Top 100 
               has changed over time. Diversity peaked in the 2000s possibly due to digital streaming 
               increasing song turnover"
  )
```

### Part C

```{r, echo = FALSE}
ten_week_hits <- billboard %>%
  group_by(performer, song) %>%
  summarize(weeks_on_chart = n(), .groups = "drop") %>%
  filter(weeks_on_chart >= 10) %>%
  group_by(performer) %>%
  summarize(num_ten_week_hits = n(), .groups = "drop") %>%
  filter(num_ten_week_hits >= 30) %>%
  arrange(desc(num_ten_week_hits))

ggplot(ten_week_hits, aes(x = reorder(performer, num_ten_week_hits), y = num_ten_week_hits)) +
  geom_bar(stat = "identity", fill = "blue") +
  coord_flip() +
  labs(
    title = "Artists with the Most 'Ten-Week Hits' on Billboard 100",
    x = "Artist",
    y = "Number of Songs with 10+ Weeks on Billboard",
    caption = "This bar plot shows the 19 artists with at least 30 songs that 
              stayed on the Billboard Top 100 for 10 or more weeks. Elton ranks at 
              the top of these artists with 52 songs 10+ on billboard"
  )
```
